Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-09 latex/main.tex
Changes for concordance table

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
recombination node (A); the recombination node associated with a group of
``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
RecombinHunt \cite{Alfonsi2024},
CovRecomb \cite{Li2024}, and
rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach:
they scan candidate recombinant sequences for mutations associated with different Pango lineages
without using a phylogeny.

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints, 
we focus on the Pango Xs associated with the Type I recombination events (Table \ref{tab:pango_x_lineages}), 
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt, we took detection results for these Pango Xs reported in the original study \cite{Alfonsi2024};
for CovRecomb, we summarized the results from the original study \cite{Li2024} (STAR Methods).
For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals 
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall, 
with comparable concordance rates (Table \ref{tab:x_method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{Concordance among methods in characterizing Pango X lineages.}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\end{tabular}
\end{table}

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated
recombination node (A); the recombination node associated with a group of
``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C).
For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

%%% TODO: Mention RIPPLES somewhere here? Or just in Discussion?
We compared sc2ts against three state-of-the-art methods that detect inter-lineage recombinants:
RecombinHunt\cite{Alfonsi2024}, CovRecomb\cite{Li2024}, and rebar (https://github.com/phac-nml/rebar).
These methods use the same underlying approach:
they scan candidate recombinant sequences for mutations associated with different Pango lineages
without using a phylogeny.

To evaluate the concordance of the methods in
identifying the Pango lineages of recombinant parents and locating recombination breakpoints,
we focus on the Pango Xs associated with the Type I recombination events (Table~\ref{tab:pango_x_lineages}),
using the data from the Pango designation community (https://github.com/cov-lineages/pango-designation/) as a common comparison.
For RecombinHunt and CovRecomb, we summarized detection results for these Pango Xs, if available,
reported in the original studies\cite{Alfonsi2024,Li2024} (STAR Methods).
For each method, we determined the Pango Xs for which the inferred parental lineages and breakpoint intervals
are concordant with those identified by the community (STAR Methods).
We found that the methods performed well overall,
with comparable concordance rates (Table~\ref{tab:method_concordance}),
indicating that sc2ts can characterize recombinants as well as the state-of-the-art methods.

@@

\begin{table}
\caption{
Concordance among methods in characterizing Pango X lineages.
Only the Pango X lineages associated with the Type 1 recombination events
are considered in these comparisons.
For each method, we calculated the number of Pango X lineages
which have inferred parent Pango lineages and breakpoint intervals
concordant with those proposed by the community.
The denominators indicate the number of Pango X lineages
for which detection results are available from the original studies.
For the Pango X lineages with non-overlapping breakpoint intervals,
the distances between the inferred breakpoint intervals and 
those proposed by the community are shown (median and range).
Related to Table \ref{tab:pango_x_lineages}.
}
\label{tab:method_concordance}
\begin{tabular}{lccc}
\toprule & \multicolumn{1}{c}{Parent lineages} & \multicolumn{2}{c}{Breakpoint intervals} \\
\cmidrule(lr){2-2} \cmidrule(lr){3-4}
Method & Concordant (\%) & Concordant (\%) & Distance (bases) \\
\midrule
Sc2ts & 14 / 15 (93.3) & 13 / 15 (86.7) & 540.5 (315, 766)\\
RecombinHunt-GISAID & 13 / 13 (100.0) & 10 / 13 (76.9) & 764.0 (4, 1202)\\
RecombinHunt-Nextstrain & 12 / 14 (85.7) & 11 / 14 (78.6) & 1160.0 (559, 2785)\\
CovRecomb & 7 / 8 (87.5) & 4 / 8 (50.0) & 1562.5 (217, 13208)\\
\bottomrule
\end{tabular}
\end{table}
